\part{Core Rules}\index{Core Rules}

\chapter{Introduction}\index{Core Rules!Introduction}

\begin{multicols}{2}

\section{How to Play the Cypher System}\index{Core Rules!Introduction!How to Play Cypher System}
The rules of the Cypher System are quite straightforward at their heart, as all of gameplay is based around a few core concepts.

This chapter provides a brief explanation of how to play the game, and it’s useful for learning the game. Once you understand the basic concepts, you’ll likely want to reference Rules of the Game for a more in-depth treatment. 

The Cypher System uses a twenty-sided die (1d20) to determine the results of most actions. Whenever a roll of any kind is called for and no die is specified, roll a d20.

The game master sets a difficulty for any given task. There are ten degrees of difficulty. Thus, the difficulty of a task can be rated on a scale of 1 to 10.

Each difficulty has a target number associated with it. The target number is always three times the task’s difficulty, so a difficulty 1 task has a target number of 3, but a difficulty 4 task has a target number of 12. To succeed at the task, you must roll the target number or higher. See the Task Difficulty table for guidance in how this works.

Character skills, favorable circumstances, or excellent equipment can decrease the difficulty of a task. For example, if a character is trained in climbing, they turn a difficulty 6 climb into a difficulty 5 climb. This is called easing the difficulty by one step (or just easing the difficulty, which assumes it’s eased by one step). If they are specialized in climbing, they turn a difficulty 6 climb into a difficulty 4 climb. This is called easing the difficulty by two steps. Decreasing the difficulty of a task can also be called easing a task. Some situations increase, or hinder, the difficulty of a task. If a task is hindered, it increases the difficulty by one step.

A skill is a category of knowledge, ability, or activity relating to a task, such as climbing, geography, or persuasiveness. A character who has a skill is better at completing related tasks than a character who lacks the skill. A character’s level of skill is either trained (reasonably skilled) or specialized (very skilled).
If you are trained in a skill relating to a task, you ease the difficulty of that task by one step. If you are specialized, you ease the difficulty by two steps. A skill can never decrease a task’s difficulty by more than two steps.

Anything else that reduces difficulty (help from an ally, a particular piece of equipment, or some other advantage) is referred to as an asset. Assets can never decrease a task’s difficulty by more than two steps.

You can also decrease the difficulty of a given task by applying Effort. (Effort is described in more detail in the Rules of the Game chapter.) 
To sum up, three things can decrease a task’s difficulty: skills, assets, and Effort.

If you can ease a task so its difficulty is reduced to 0, you automatically succeed and don’t need to make a roll.

\subsection{When Do You Roll?}\index{Core Rules!Introduction!When Do You Roll?} 

Any time your character attempts a task, the GM assigns a difficulty to that task, and you roll a d20 against the associated target number.

When you jump from a burning vehicle, swing an axe at a mutant beast, swim across a raging river, identify a strange device, convince a merchant to give you a lower price, craft an object, use a power to control a foe’s mind, or use a blaster rifle to carve a hole in a wall, you make a d20 roll.

However, if you attempt something that has a difficulty of 0, no roll is needed—you automatically succeed. Many actions have a difficulty of 0. Examples include walking across the room and opening a door, using a special ability to negate gravity so you can fly, using an ability to protect your friend from radiation, or activating a device (that you already understand) to erect a force field. These are all routine actions and don’t require rolls.

Using skill, assets, and Effort, you can ease the difficulty of potentially any task to 0 and thus negate the need for a roll. Walking across a narrow wooden beam is tricky for most people, but for an experienced gymnast, it’s routine. You can even ease the difficulty of an attack on a foe to 0 and succeed without rolling.

If there’s no roll, there’s no chance for failure. However, there’s also no chance for remarkable success (in the Cypher System, that usually means rolling a 19 or 20, which are called special rolls; the Rules of the Game chapter also discusses special rolls).

\subsection{Task Difficulty}

\end{multicols}

\begin{table}
\centering

\caption{Task Difficulty}
\label{tab:Task Difficulty}

\begin{tabularx}{\textwidth}{| p{0.15\textwidth} | p{0.3\textwidth} | p{0.15\textwidth} | p{0.3\textwidth} |}
\hline
\textbf{Task \newline Difficulty} & \textbf{Description} & \textbf{Target Number} & \textbf{Guidance} \\
\hline
% Tutaj wpisujesz dane do tabeli
0 & Routine & 0 & Anyone can do this basically every time.  \\ \hline
1 & Simple & 3 & Most people can do this most of the time.  \\ \hline
2 & Standard & 6 & Typical task requiring focus, but most people can usually do this.  \\ \hline
3 & Demanding & 9 & Requires full attention; most people have a 50/50 chance to succeed.  \\ \hline
4 & Difficult & 12 & Trained people have a 50/50 chance to succeed.  \\ \hline
5 & Challenging & 15 & Even trained people often fail.  \\ \hline
6 & Intimidating & 18 & Normal people almost never succeed.  \\ \hline
7 & Formidable & 21 & Impossible without skills or great effort.  \\ \hline
8 & Heroic & 24 & A task worthy of tales told for years afterward.  \\ \hline
9 & Immortal & 27 & A task worthy of legends that last lifetimes.  \\ \hline
10 & Impossible & 30 & A task that normal humans couldn’t consider (but one that doesn’t break the laws of physics).  \\ \hline

\end{tabularx}

\end{table}

\begin{multicols}{2}

\subsection{Combat}\index{Core Rules!Introduction!Combat}

Making an attack in combat works the same way as any other roll: the GM assigns a difficulty to the task, and you roll a d20 against the associated target number.

The difficulty of your attack roll depends on how powerful your opponent is. Just as tasks have a difficulty from 1 to 10, creatures have a level from 1 to 10. Most of the time, the difficulty of your attack roll is the same as the creature’s level. For example, if you attack a level 2 bandit, it’s a level 2 task, so your target number is 6.

It’s worth noting that players make all die rolls. If a character attacks a creature, the player makes an attack roll. If a creature attacks a character, the player makes a defense roll.

The damage dealt by an attack is not determined by a roll—it’s a flat number based on the weapon or attack used. For example, a spear always does 4 points of damage.

Your Armor characteristic reduces the damage you take from attacks directed at you. You get Armor from wearing physical armor (such as a leather jacket in a modern game or chainmail in a fantasy setting) or from special abilities. Like weapon damage, Armor is a flat number, not a roll. If you’re attacked, subtract your Armor from the damage you take. For example, a leather jacket gives you +1 to Armor, meaning that you take 1 less point of damage from attacks. If a mugger hits you with a knife for 2 points of damage while you’re wearing a leather jacket, you take only 1 point of damage. If your Armor reduces the damage from an attack to 0, you take no damage from that attack.

When you see the word “Armor” capitalized in the game rules (other than in the name of a special ability), it refers to your Armor characteristic—the number you subtract from incoming damage. When you see the word “armor” with a lowercase “a,” it refers to any physical armor you might wear.
Typical physical weapons come in three categories: light, medium and heavy.

Light weapons inflict only 2 points of damage, but they ease attack rolls because they are fast and easy to use. Light weapons are punches, kicks, clubs, knives, handaxes, rapiers, small pistols, and so on. Weapons that are particularly small are light weapons.

Medium weapons inflict 4 points of damage. Medium weapons include swords, battleaxes, maces, crossbows, spears, pistols, blasters, and so on. Most weapons are medium. Anything that could be used in one hand (even if it’s often used in two hands, such as a quarterstaff or spear) is a medium weapon.

Heavy weapons inflict 6 points of damage, and you must use two hands to attack with them. Heavy weapons are huge swords, great hammers, massive axes, halberds, heavy crossbows, blaster rifles, and so on. Anything that must be used in two hands is a heavy weapon.

\subsection{Special Rolls}\index{Core Rules!Introduction!Special Rolls}

When you roll a natural 19 (the d20 shows “19”) and the roll is a success, you also have a minor effect. In combat, a minor effect inflicts 3 additional points of damage with your attack, or, if you’d prefer a special result, you could decide instead that you knock the foe back, distract them, or something similar. When not in combat, a minor effect could mean that you perform the action with particular grace. For example, when jumping down from a ledge, you land smoothly on your feet, or when trying to persuade someone, you convince them that you’re smarter than you really are. In other words, you not only succeed but also go a bit further.

When you roll a natural 20 (the d20 shows “20”) and the roll is a success, you also have a major effect. This is similar to a minor effect, but the results are more remarkable. In combat, a major effect inflicts 4 additional points of damage with your attack, but again, you can choose instead to introduce a dramatic event such as knocking down your foe, stunning them, or taking an extra action. Outside of combat, a major effect means that something beneficial happens based on the circumstance. For example, when climbing up a cliff wall, you make the ascent twice as fast. When a roll grants you a major effect, you can choose to use a minor effect instead if you prefer.

In combat (and only in combat), if you roll a natural 17 or 18 on your attack roll, you add 1 or 2 additional points of damage, respectively. Neither roll has any special effect options—just the extra damage.

(For more information on special rolls and how they affect combat and other interactions, see Rules of the Game.)

Rolling a natural 1 is always bad. It means that the GM introduces a new complication into the encounter.

\subsection{Glossary}\index{Core Rules!Introduction!Glossary}

\textbf{Game master (GM)}: The player who doesn’t run a character, but instead guides the flow of the story and runs all the NPCs.

\textbf{Nonplayer character (NPC)}: Characters run by the GM. Think of them as the minor characters in the story, or the villains or opponents. This includes any kind of creature as well as people.

\textbf{Party}: A group of player characters (and perhaps some NPC allies).

\textbf{Player character (PC)}: A character run by a player rather than the GM. Think of the PCs as the main characters in the story.

\textbf{Player}: The players who run characters in the game.

\textbf{Session}: A single play experience. Usually lasts a few hours. Sometimes one adventure can be accomplished in a session. More often, one adventure is multiple sessions.

\textbf{Adventure}: A single portion of the campaign with a beginning and an end. Usually defined at the beginning by a goal put forth by the PCs and at the end by whether or not they achieve that goal.

\textbf{Campaign}: A series of sessions strung together with an overarching story (or linked stories) with the same player characters. Often, but not always, a campaign involves a number of adventures.

\textbf{Character}: Anything that can act in the game. Although this includes PCs and human NPCs, it also technically includes creatures, aliens, mutants, automatons, animate plants, and so on. The word “creature” is usually synonymous.

\subsection{Range and speed}\index{Core Rules!Introduction!Range and Speed}

Distance is simplified into four categories: immediate, short, long, and very long.

Immediate distance from a character is within reach or within a few steps. If a character stands in a small room, everything in the room is within immediate distance. At most, immediate distance is 10 feet (3 m).

Short distance is anything greater than immediate distance but less than 50 feet (15 m) or so. 

Long distance is anything greater than short distance but less than 100 feet (30 m) or so. 

Very long distance is anything greater than long distance but less than 500 feet (150 m) or so. Beyond that range, distances are always specified—1,000 feet (300 m), a mile (1.5 km), and so on.

The idea is that it’s not necessary to measure precise distances. Immediate distance is right there, practically next to the character. Short distance is nearby. Long distance is farther off. Very long distance is really far off.

All weapons and special abilities use these terms for ranges. For example, all melee weapons have immediate range—they are close-combat weapons, and you can use them to attack anyone within immediate distance. A thrown knife (and most other thrown weapons) has short range. A bow has long range. An Adept’s Onslaught ability also has short range.

A character can move an immediate distance as part of another action. In other words, they can take a few steps over to the control panel and activate a switch. They can lunge across a small room to attack a foe. They can open a door and step through.

A character can move a short distance as their entire action for a turn. They can also try to move a long distance as their entire action, but the player might have to roll to see if the character slips, trips, or stumbles as the result of moving so far so quickly.

For example, if the PCs are fighting a group of cultists, any character can likely attack any cultist in the general melee—they’re all within immediate range. Exact positions aren’t important. Creatures in a fight are always moving, shifting, and jostling, anyway. However, if one cultist stayed back to fire a pistol, a character might have to use their entire action to move the short distance required to attack that foe. It doesn’t matter if the cultist is 20 feet (6 m) or 40 feet (12 m) away—it’s simply considered short distance. It does matter if the cultist is more than 50 feet (15 m) away because that distance would require a long or very long move.

(Many rules in this system avoid the cumbersome need for precision. Does it really matter if the ghost is 13 feet away from you or 18? Probably not. That kind of needless specificity only slows things down and draws away from, rather than contributes to, the story.)

\subsection{Experience Points}\index{Core Rules!Introduction!Experience Points}

Experience points (XP) are rewards given to players when the GM intrudes on the story (this is called GM intrusion) with a new and unexpected challenge. For example, in the middle of combat, the GM might inform the player that they drop their weapon. However, to intrude in this manner, the GM must award the player 2 XP. The rewarded player, in turn, must immediately give one of those XP to another player and justify the gift (perhaps the other player had a good idea, told a funny joke, performed an action that saved a life, and so on).

Alternatively, the player can refuse the GM intrusion. If they do so, they don’t get the 2 XP from the GM, and they must also spend 1 XP that they already have. If the player has no XP to spend, they can’t refuse the intrusion.

The GM can also give players XP between sessions as a reward for making discoveries during an adventure. Discoveries are interesting facts, wondrous secrets, powerful artifacts, answers to mysteries, or solutions to problems (such as where the kidnappers are keeping their victim or how the PCs repair the starship). You don’t earn XP for killing foes or overcoming standard challenges in the course of play. Discovery is the soul of the Cypher System.

Experience points are used primarily for character advancement (for details, see the Creating Your Character chapter), but a player can also spend 1 XP to reroll any die roll and take the better of the two rolls.

\subsection{Cyphers}\index{Core Rules!Introduction!Cyphers}

Cyphers are abilities that have a single use. In many campaigns, cyphers aren’t physical objects—they might be a spell cast upon a character, a blessing from a god, or just a quirk of fate that gives them a momentary advantage. In some campaigns, cyphers are physical objects that characters can carry. Whether or not cyphers are physical objects, they are part of the character (like equipment or a special ability) and are things characters can use during the game. The form that physical cyphers take depends on the setting. In a fantasy world they might be wands or potions, but in a science fiction game they could be alien crystals or prototype devices.

Characters will find new cyphers frequently in the course of play, so players shouldn’t hesitate to use their cypher abilities. Because cyphers are always different, the characters will always have new special powers to try.

\subsection{Other Dice}\index{Core Rules!Introduction!Other Dice}

In addition to a d20, you’ll need a d6 (a six-sided die). Rarely, you’ll need to roll a number between 1 and 100 (often called a d100 or d\% roll), which you can do by rolling a d20 twice, using the last digit of the first roll as the “tens” place and the last digit of the second roll as the “ones” place. For example, rolling a 17 and a 9 gives you 79, rolling a 3 and an 18 gives you 38, and rolling a 20 and a 10 gives you 00 (also known as 100). If you have a d10 (a ten-sided die), you can use it instead of the d20 to roll numbers between 1 and 100.

(A d6 is used most often for recovery rolls and to determine the level of cyphers.

\subsection{Creating Your Character}\index{Core Rules!Introduction!Creating Your Character}

This section explains how to create characters to play in a Cypher System game. This involves a series of decisions that will shape your character, so the more you understand what kind of character you want to play, the easier character creation will be. The process involves understanding the values of three game statistics and choosing three aspects that determine your character’s capabilities.

\subsubsection{Character Stats}\index{Core Rules!Introduction!Character Stats}

Every player character has three defining characteristics, which are typically called “statistics” or “stats.” These stats are Might, Speed, and Intellect. They are broad categories that cover many different but related aspects of a character.

\paragraph{Might}\index{Core Rules!Introduction!Character Stats!Might}

Might defines how strong and durable your character is. The concepts of strength, endurance, constitution, hardiness, and physical prowess are all folded into this one stat. Might isn’t relative to size; instead, it’s an absolute measurement. An elephant has more Might than the mightiest tiger, which has more Might than the mightiest rat, which has more Might than the mightiest spider.

Might governs actions from forcing doors open to walking for days without food to resisting disease. It’s also the primary means of determining how much damage your character can sustain in a dangerous situation. Physical characters, tough characters, and characters interested in fighting should focus on Might.

(Might could be thought of as Might/Health because it governs how strong you are and how much physical punishment you can take.)

\paragraph{Speed}\index{Core Rules!Introduction!Character Stats!Speed}

Speed describes how fast and physically coordinated your character is. The stat embodies quickness, movement, dexterity, and reflexes. Speed governs such divergent actions as dodging attacks, sneaking around quietly, and throwing a ball accurately. It helps determine whether you can move farther on your turn. Nimble, fast, or sneaky characters will want good Speed stats, as will those interested in ranged combat.

(Speed could be thought of as Speed/Agility because it governs your overall swiftness and reflexes.)

\paragraph{Intelect}\index{Core Rules!Introduction!Character Stats!Intelect}

This stat determines how smart, knowledgeable, and likable your character is. It includes intelligence, wisdom, charisma, education, reasoning, wit, willpower, and charm. Intellect governs solving puzzles, remembering facts, telling convincing lies, and using mental powers. Characters interested in communicating effectively, being learned scholars, or wielding supernatural powers should stress their Intellect stat.

(Intellect could be thought of as Intellect/Personality because it governs both intelligence and charisma.)

\subsection{Pool, Edge and Effort}\index{Core Rules!Introduction!Character Stats!Pool, Edge and Effort}

Each of the three stats has two components: Pool and Edge. Your Pool represents your raw, innate ability, and your Edge represents knowing how to use what you have. A third element ties into this concept: Effort. When your character really needs to accomplish a task, you apply Effort.

(Your stat Pools, as well as your Effort and Edge, are determined by the character type, descriptor, and focus that you choose. Within those guidelines, however, you have a lot of flexibility in how you develop your character.)

\paragraph{Pool}\index{Core Rules!Introduction!Character Stats!Pool, Edge and Effort!Pool}

Your Pool is the most basic measurement of a stat. Comparing the Pools of two creatures will give you a general sense of which creature is superior in that stat. For example, a character who has a Might Pool of 16 is stronger (in a basic sense) than a character who has a Might Pool of 12. Most characters start with a Pool of 9 to 12 in most stats—that’s the average range.

When your character is injured, sickened, or attacked, you temporarily lose points from one of your stat Pools. The nature of the attack determines which Pool loses points. For example, physical damage from a sword reduces your Might Pool, a poison that makes you clumsy reduces your Speed Pool, and a psionic blast reduces your Intellect Pool. You can also spend points from one of your stat Pools to decrease a task’s difficulty (see Effort, below). You can rest to recover lost points from a stat Pool, and some special abilities or cyphers might allow you to recover lost points quickly.

\paragraph{Edge}\index{Core Rules!Introduction!Character Stats!Pool, Edge and Effort!Edge}

Although your Pool is the basic measurement of a stat, your Edge is also important. When something requires you to spend points from a stat Pool, your Edge for that stat reduces the cost. It also reduces the cost of applying Effort to a roll.

For example, let’s say you have a mental blast ability, and activating it costs 1 point from your Intellect Pool. Subtract your Intellect Edge from the activation cost, and the result is how many points you must spend to use the mental blast. If using your Edge reduces the cost to 0, you can use the ability for free.

Your Edge can be different for each stat. For example, you could have a Might Edge of 1, a Speed Edge of 1, and an Intellect Edge of 0. You’ll always have an Edge of at least 1 in one stat. Your Edge for a stat reduces the cost of spending points from that stat Pool, but not from other Pools. Your Might Edge reduces the cost of spending points from your Might Pool, but it doesn’t affect your Speed Pool or Intellect Pool. Once a stat’s Edge reaches 3, you can apply one level of Effort for free.

A character who has a low Might Pool but a high Might Edge has the potential to perform Might actions consistently better than a character who has a Might Edge of 0. The high Edge will let them reduce the cost of spending points from the Pool, which means they’ll have more points available to spend on applying Effort.

\paragraph{Effort}\index{Core Rules!Introduction!Character Stats!Pool, Edge and Effort!Effort}

When your character really needs to accomplish a task, you can apply Effort. For a beginning character, applying Effort requires spending 3 points from the stat Pool appropriate to the action. Thus, if your character tries to dodge an attack (a Speed roll) and wants to increase the chance for success, you can apply Effort by spending 3 points from your Speed Pool. Effort eases the task by one step. This is called applying one level of Effort.

You don’t have to apply Effort if you don’t want to. If you choose to apply Effort to a task, you must do it before you attempt the roll—you can’t roll first and then decide to apply Effort if you rolled poorly.

Applying more Effort can lower a task’s difficulty further: each additional level of Effort eases the task by another step. Applying one level of Effort eases the task by one step, applying two levels eases the task by two steps, and so on. However, each level of Effort after the first costs only 2 points from the stat Pool instead of 3. So applying two levels of Effort costs 5 points (3 for the first level plus 2 for the second level), applying three levels costs 7 points (3 plus 2 plus 2), and so on.

Every character has an Effort score, which indicates the maximum number of levels of Effort that can be applied to a roll. A beginning (first-tier) character has an Effort of 1, meaning you can apply only one level of Effort to a roll. A more experienced character has a higher Effort score and can apply more levels of Effort to a roll. For example, a character who has an Effort of 3 can apply up to three levels of Effort to reduce a task’s difficulty.

When you apply Effort, subtract your relevant Edge from the total cost of applying Effort. For example, let’s say you need to make a Speed roll. To increase your chance for success, you decide to apply one level of Effort, which will ease the task. Normally, that would cost 3 points from your Speed Pool. However, you have a Speed Edge of 2, so you subtract that from the cost. Thus, applying Effort to the roll costs only 1 point from your Speed Pool.

What if you applied two levels of Effort to the Speed roll instead of just one? That would ease the task by two steps. Normally, it would cost 5 points from your Speed Pool, but after subtracting your Speed Edge of 2, it costs only 3 points.

Once a stat’s Edge reaches 3, you can apply one level of Effort for free. For example, if you have a Speed Edge of 3 and you apply one level of Effort to a Speed roll, it costs you 0 points from your Speed Pool. (Normally, applying one level of Effort would cost 3 points, but you subtract your Speed Edge from that cost, reducing it to 0.)

Skills and other advantages also ease a task, and you can use them in conjunction with Effort. In addition, your character might have special abilities or equipment that allow you to apply Effort to accomplish a special effect, such as knocking down a foe with an attack or affecting multiple targets with a power that normally affects only one.

(When applying Effort to melee attacks, you have the option of spending points from either your Might Pool or your Speed Pool. When making ranged attacks, you may spend points only from your Speed Pool. This reflects that with melee you sometimes use brute force and sometimes use finesse, but with ranged attacks, it’s always about careful targeting.)

\paragraph{Effort and Damage}\index{Core Rules!Introduction!Character Stats!Pool, Edge and Effort!Effort and Damage}

nstead of applying Effort to ease your attack, you can apply Effort to increase the amount of damage you inflict with an attack. For each level of Effort you apply in this way, you inflict 3 additional points of damage. This works for any kind of attack that inflicts damage, whether a sword, a crossbow, a mind blast, or something else.

When using Effort to increase the damage of an area attack, such as the explosion created by an Adept’s Concussion ability, you inflict 2 additional points of damage instead of 3 points. However, the additional points are dealt to all targets in the area. Further, even if one or more of the targets resist the attack, they still take 1 point of damage.

\paragraph{Multiple Uses of Effort and Edge}\index{Core Rules!Introduction!Character Stats!Pool, Edge and Effort!Multiple Uses of Effort and Edge}

f your Effort is 2 or higher, you can apply Effort to multiple aspects of a single action. For example, if you make an attack, you can apply Effort to your attack roll and apply Effort to increase the damage.

The total amount of Effort you apply can’t be higher than your Effort score. For example, if your Effort is 2, you can apply up to two levels of Effort. You could apply one level to an attack roll and one level to its damage, two levels to the attack and no levels to the damage, or no levels to the attack and two levels to the damage.

You can use Edge for a particular stat only once per action. For example, if you apply Effort to a Might attack roll and to your damage, you can use your Might Edge to reduce the cost of one of those uses of Effort, not both. If you spend 1 Intellect point to activate your mind blast and one level of Effort to ease the attack roll, you can use your Intellect Edge to reduce the cost of one of those things, not both.

\subsubsection{Character Stats}\index{Core Rules!Introduction!Character Stats!Stats Examples}

A beginning character is fighting a giant rat. The PC stabs their spear at the rat, which is a level 2 creature and thus has a target number of 6. The character stands atop a boulder and strikes downward at the beast, and the GM rules that this helpful tactic is an asset that eases the attack by one step (to difficulty 1). That lowers the target number to 3. Attacking with a spear is a Might action; the character has a Might Pool of 11 and a Might Edge of 0. Before making the roll, they decide to apply a level of Effort to ease the attack. That costs 3 points from their Might Pool, reducing the Pool to 8. But the points are well spent. Applying the Effort lowers the difficulty from 1 to 0, so no roll is needed—the attack automatically succeeds.

Another character is attempting to convince a guard to let them into a private office to speak to an influential noble. The GM rules that this is an Intellect action. The character is third tier and has an Effort of 3, an Intellect Pool of 13, and an Intellect Edge of 1. Before making the roll, they must decide whether to apply Effort. They can choose to apply one, two, or three levels of Effort, or apply none at all. This action is important to them, so they decide to apply two levels of Effort, easing the task by two steps. Thanks to their Intellect Edge, applying the Effort costs only 4 points from their Intellect Pool (3 points for the first level of Effort plus 2 points for the second level minus 1 point for their Edge). Spending those points reduces their Intellect Pool to 9. The GM decides that convincing the guard is a difficulty 3 (demanding) task with a target number of 9; applying two levels of Effort reduces the difficulty to 1 (simple) and the target number to 3. The player rolls a d20 and gets an 8. Because this result is at least equal to the target number of the task, they succeed. However, if they had not applied some Effort, they would have failed because their roll (8) would have been less than the task’s original target number (9).

\subsection{Character Tiers}\index{Core Rules!Introduction!Character Tiers}

Every character starts the game at the first tier. Tier is a measurement of power, toughness, and ability. Characters can advance up to the sixth tier. As your character advances to higher tiers, you gain more abilities, increase your Effort, and can improve a stat’s Edge or increase a stat. Generally speaking, even first-tier characters are already quite capable. It’s safe to assume that they’ve already got some experience under their belt. This is not a “zero to hero” progression, but rather an instance of competent people refining and honing their capabilities and knowledge. Advancing to higher tiers is not really the goal of Cypher System characters, but rather a representation of how characters progress in a story.

To progress to the next tier, characters earn experience points (XP) by pursuing character arcs, going on adventures, and discovering new things—the system is about both discovery and exploration, as well as achieving personal goals. Experience points have many uses, and one use is to purchase character benefits. After your character purchases four character benefits, they advance to the next tier. Each benefit costs 4 XP, and you can purchase them in any order, but you must purchase one of each kind of benefit (and then advance to the next tier) before you can purchase the same benefit again. The four character benefits are as follows.

\textbf{Increasing Capabilities}: You gain 4 points to add to your stat Pools. You can allocate the points among the Pools however you wish.

\textbf{Moving Toward Perfection}: You add 1 to your Might Edge, your Speed Edge, or your Intellect Edge (your choice).

\textbf{Extra Effort}: Your Effort score increases by 1.

\textbf{Skills}: You become trained in one skill of your choice, other than attacks or defense. As described in Rules of the Game, a character trained in a skill treats the difficulty of a related task as one step lower than normal. The skill you choose for this benefit can be anything you wish, such as climbing, jumping, persuading, or sneaking. You can also choose to be knowledgeable in a certain area of lore, such as history or geology. You can even choose a skill based on your character’s special abilities. For example, if your character can make an Intellect roll to blast an enemy with mental force, you can become trained in using that ability, easing the task of using it. If you choose a skill that you are already trained in, you become specialized in that skill, easing related tasks by two steps instead of one.

(Skills are a broad category of things your character can learn and accomplish. For a list of sample skills, see below.)

\textbf{Other Options}: Players can also spend 4 XP to purchase other special options in lieu of gaining a new skill. Selecting any of these options counts as the skill benefit necessary to advance to the next tier. The special options are as follows:

\begin{itemize}
\item Reduce the cost for wearing armor. This option lowers the Speed cost for wearing armor by 1. 
\item Add 2 to your recovery rolls.
\item Select a new type-based ability from your tier or a lower tier.
\end{itemize}

\subsection{Character Descriptor, Type, and Focus}\index{Core Rules!Introduction!Character Descriptor, Type and Focus}

To create your character, you build a simple statement that describes them. The statement takes this form: “I am a [fill in an adjective here] [fill in a noun here] who [fill in a verb here].”

Thus: “I am an adjective noun who verbs.” For example, you might say, “I am a Rugged Warrior who Controls Beasts” or “I am a Charming Explorer who Focuses Mind Over Matter.”

In this sentence, the adjective is called your descriptor.

The noun is your character type.

The verb is called your focus.

Even though character type is in the middle of the sentence, that’s where we’ll start this discussion. (Just as in a sentence, the noun provides the foundation.)
Your character type is the core of your character. In some roleplaying games, it might be called your character class. Your type helps determine your character’s place in the world and relationship with other people in the setting. It’s the noun of the sentence “I am an adjective noun who verbs.”

You can choose from four character types: Warriors, Adepts, Explorers, and Speakers.

Your descriptor defines your character—it colors everything you do. Your descriptor places your character in the situation (the first adventure, which starts the campaign) and helps provide motivation. It’s the adjective of the sentence “I am an adjective noun who verbs.”

Unless your GM says otherwise, you can choose from any of the character descriptors.

Focus is what your character does best. Focus gives your character specificity and provides interesting new abilities that might come in handy. Your focus also helps you understand how you relate with the other player characters in your group. It’s the verb of the sentence “I am an adjective noun who verbs.”
There are many character foci. The ones you choose from will probably depend on the setting and genre of your game.

(You can use the Flavors chapter to slightly modify character types to customize them for different genres.)

\subsection{Special Abilities}\index{Core Rules!Introduction!Special Abilities}

Character types and foci grant PCs special abilities at each new tier. Using these abilities usually costs points from your stat Pools; the cost is listed in parentheses after the ability name. Your Edge in the appropriate stat can reduce the cost of the ability, but remember that you can apply Edge only once per action. For example, let’s say an Adept with an Intellect Edge of 2 wants to use their Onslaught ability to create a bolt of force, which costs 1 Intellect point. They also want to increase the damage from the attack by using a level of Effort, which costs 3 Intellect points. The total cost for their action is 2 points from their Intellect Pool (1 point for the bolt of force, plus 3 points for using Effort, minus 2 points from their Edge).

Sometimes the point cost for an ability has a + sign after the number. For example, the cost might be given as “2+ Intellect points.” That means you can spend more points or more levels of Effort to improve the ability further, as explained in the ability description.

Many special abilities grant a character the option to perform an action that they couldn’t normally do, such as projecting rays of cold or attacking multiple foes at once. Using one of these abilities is an action unto itself, and the end of the ability’s description says “Action” to remind you. It also might provide more information about when or how you perform the action.

Some special abilities allow you to perform a familiar action—one that you can already do—in a different way. For example, an ability might let you wear heavy armor, reduce the difficulty of Speed defense rolls, or add 2 points of fire damage to your weapon damage. These abilities are called enablers. Using one of these abilities is not considered an action. Enablers either function constantly (such as being able to wear heavy armor, which isn’t an action) or happen as part of another action (such as adding fire damage to your weapon damage, which happens as part of your attack action). If a special ability is an enabler, the end of the ability’s description says “Enabler” to remind you.

Some abilities specify a duration, but you can always end one of your own abilities anytime you wish.

(Because the Cypher System covers so many genres, not all of the descriptors, types, and foci might be available for players. The GM will decide what’s available in their particular game and whether anything is modified, and they’ll let the players know.)

\subsection{Skills}\index{Core Rules!Introduction!Skills}

Sometimes your character gains training in a specific skill or task. For example, your focus might mean that you’re trained in sneaking, in climbing and jumping, or in social interactions. Other times, your character can choose a skill to become trained in, and you can pick a skill that relates to any task you think you might face.

The Cypher System has no definitive list of skills. However, the following list offers ideas:

\begin{itemize}
\item Astronomy
\item Balancing
\item Biology
\item Botany
\item Carrying
\item Climbing
\item Computers
\item Deceiving
\item Disguise
\item Escaping
\item Geography
\item Geology
\item Healing
\item History
\item Identifying
\item Initiative
\item Intimidation
\item Jumping
\item Leatherworking
\item Lockpicking
\item Machinery
\item Metalworking
\item Perception
\item Persuasion
\item Philosophy
\item Physics
\item Pickpocketing
\item Piloting
\item Repairing
\item Riding
\item Smashing
\item Sneaking
\item Stealth
\item Swimming
\item Vehicle driving
\item Woodworking 
\end{itemize}

You could choose a skill that incorporates more than one of these areas (interacting might include deceiving, intimidation, and persuasion) or that is a more specific version of one (hiding might be sneaking when you’re not moving). You could also make up more general professional skills, such as baker, sailor, or lumberjack. If you want to choose a skill that’s not on this list, it’s probably best to run it past the GM first, but in general, the most important thing is to choose skills that are appropriate to your character.

Remember that if you gain a skill that you’re already trained in, you become specialized in that skill. Because skill descriptions can be nebulous, determining whether you’re trained or specialized might take some thinking. For example, if you’re trained in lying and later gain an ability that grants you skill with all social interactions, you become specialized in lying and trained in all other types of interactions. Being trained three times in a skill is no better than being trained twice (in other words, specialized is as good as it gets).

Only skills gained through character type abilities or other rare instances allow you to become skilled with attack or defense tasks.

If you gain a special ability through your type, your focus, or some other aspect of your character, you can choose it in place of a skill and become trained or specialized in that ability. For example, if you have a mind blast, when it’s time to choose a skill to be trained in, you can select your mind blast as your skill. That would ease the attack every time you used it. Each ability you have counts as a separate skill for this purpose. You can’t select “all mind powers” or “all spells” as one skill and become trained or specialized in such a broad category.

In most campaigns, fluency in a language is considered a skill. So if you want to speak French, that’s the same as being trained in biology or swimming.

\subsection{Type}\index{Core Rules!Introduction!Type}

Character type is the core of your character. Your type helps determine your character’s place in the world and relationship with other people in the setting. It’s the noun of the sentence “I am an adjective noun who verbs.”

(In some roleplaying games, your character type might be called your character class.)

You can choose from four character types: Warrior, Adept, Explorer, and Speaker. However, you may not want to use these generic names for them. This chapter offers a few more specific names for each type that might be more appropriate to various genres. You’ll find that names like “Warrior” or “Explorer” don’t always feel right, particularly in games set in modern times. As always, you’re free to do as you wish. (Your type is who your character is. You should use whatever name you want for your type, as long as it fits both your character and the setting.)

Since the type is the basis upon which your whole character is built, it’s important to consider how the type relates to the chosen setting. To help with this, types are actually general archetypes. A Warrior, for example, might be anyone from a knight in shining armor to a cop on the streets to a grizzled cybernetic veteran of a thousand futuristic wars. 

To further massage the four types for better use in various settings, different methods called flavors are presented in Flavors to help slightly tailor the types toward fantasy, science fiction, or other genres (or to address different character concepts).

Finally, more fundamental options for further customization are provided at the end of this chapter.

\subsection{Player Intrusion}\index{Core Rules!Introduction!Player Intrusion}

A player intrusion is the player choosing to alter something in the campaign, making things easier for a player character. Conceptually, it is the reverse of a GM intrusion: instead of the GM giving the player XP and introducing an unexpected complication for a character, the player spends 1 XP and presents a solution to a problem or complication. What a player intrusion can do usually introduces a change to the world or current circumstances rather than directly changing the character. For instance, an intrusion indicating that the cypher just used still has an additional use would be appropriate, but an intrusion that heals the character would not. If a player has no XP to spend, they can’t use a player intrusion.

A few player intrusion examples are provided under each type. That said, not every player intrusion listed there is appropriate for all situations. The GM may allow players to come up with other player intrusion suggestions, but the GM is the final arbiter of whether the suggested intrusion is appropriate for the character’s type and suitable for the situation. If the GM refuses the intrusion, the player doesn’t spend the 1 XP, and the intrusion doesn’t occur.

Using an intrusion does not require a character to use an action to trigger it. A player intrusion just happens.

(Player intrusions should be limited to no more than one per player per session.)

\subsection{Defense Task}\index{Core Rules!Introduction!Defense Task}

Defense tasks are when a player makes a roll to keep something undesirable from happening to their PC. The type of defense task matters when using Effort.

\textbf{Might defense}: Used for resisting poison, disease, and anything else that can be overcome with strength and health.

\textbf{Speed defense}: Used for dodging attacks and escaping danger. This is by far the most commonly used defense task.

\textbf{Intellect defense}: Used for fending off mental attacks or anything that might affect or influence one’s mind.

\chapter{Types}\index{Types}

\section{Warrior}\index{Types!Warrior}

\textbf{Fantasy/Fairy tale}: Warrior, fighter, swordsman, knight, barbarian, soldier, myrmidon, valkyrie

\textbf{Modern/Horror/Romance}: police officer, soldier, watchman, detective, guard, brawler, tough, athlete

\textbf{Science fiction}: security officer, warrior, trooper, soldier, merc

\textbf{Superhero/Post-Apocalyptic}: hero, brick, bruiser

You’re a good ally to have in a fight. You know how to use weapons and defend yourself. Depending on the genre and setting in question, this might mean wielding a sword and shield in the gladiatorial arena, an AK-47 and a bandolier of grenades in a savage firefight, or a blaster rifle and powered armor when exploring an alien planet.

\textbf{Individual Role}: Warriors are physical, action-oriented people. They’re more likely to overcome a challenge using force than by other means, and they often take the most straightforward path toward their goals. 

\textbf{Group Role}: Warriors usually take and deal the most punishment in a dangerous situation. Often it falls on them to protect the other group members from threats. This sometimes means that warriors take on leadership roles as well, at least in combat and other times of danger.

\textbf{Societal Role}: Warriors aren’t always soldiers or mercenaries. Anyone who is ready for violence, or even potential violence, might be a Warrior in the general sense. This includes guards, watchmen, police officers, sailors, or people in other roles or professions who know how to defend themselves with skill. 

\textbf{Advanced Warriors}: As warriors advance, their skill in battle—whether defending themselves or dishing out damage—increases to impressive levels. At higher tiers, they can often take on groups of foes by themselves or stand toe to toe with anyone.

\subsection{Warrior Stat Pools}\index{Types!Warrior!Stat Pools}

\begin{table}[H]
\centering
\caption{Warrior Stat Pools}
\label{tab: Warrior Stat Pools}

\begin{tabularx}{\linewidth}{| X | X |}
% \begin{tabularx}{\linewidth}{| p{0.5\linewidth} | p{0.5\linewidth} |}
\hline
\textbf{Stat} & \textbf{Pool Starting Value} \\
\hline
% Tutaj wpisujesz dane do tabeli
Might & 10 \\ \hline
Speed & 10 \\ \hline
Intelect & 8  \\ \hline

\end{tabularx}

\end{table}

\raggedright

You get 6 additional points to divide among your stat Pools however you wish.

\subsection{Warrior Background Connection}\index{Types!Warrior!Background Connection}

Your type helps determine the connection you have to the setting. Roll a d20 or choose from the following list to determine a specific fact about your background that provides a connection to the rest of the world. You can also create your own fact.

\begin{table*}
\centering

\caption{Warrior Background Connection}
\label{tab:Warrior Background Connection}

\begin{tabularx}{\textwidth}{| p{0.05\textwidth} | p{0.9\textwidth} |}
\hline
\textbf{d20} & \textbf{Background} \\
\hline
% Tutaj wpisujesz dane do tabeli
1 & You were in the military and have friends who still serve. Your former commander remembers you well.\\ \hline
2 & You were the bodyguard of a wealthy woman who accused you of theft. You left her service in disgrace.  \\ \hline
3 & You were the bouncer in a local bar for a while, and the patrons there remember you.  \\ \hline
4 & You trained with a highly respected mentor. They regard you well, but they have many enemies.  \\ \hline
5 & You trained in an isolated monastery. The monks think of you as a brother, but you’re a stranger to all others.  \\ \hline
6 & You have no formal training. Your abilities come to you naturally (or unnaturally).  \\ \hline
7 & You spent time on the streets and were in prison for a while. \\ \hline
8 & You were conscripted into military service, but you deserted before long.  \\ \hline
9 & You served as a bodyguard to a powerful criminal who now owes you their life.  \\ \hline
10 & You worked as a police officer or constable of some kind. Everyone knows you, but their opinions of you vary.  \\ \hline
11 & Your older sibling is an infamous character who has been disgraced. \\ \hline
12 & You served as a guard for someone who traveled extensively. You know a smattering of people in many locations. \\ \hline
13 & Your best friend is a teacher or scholar. They are a great source of knowledge. \\ \hline
14 & You and a friend both smoke the same kind of rare, expensive tobacco. The two of you get together weekly to chat and smoke. \\ \hline
15 & Your uncle runs a theater in town. You know all the actors and watch all the shows for free. \\ \hline
16 & Your craftsman friend sometimes calls on you for help. However, they pay you well. \\ \hline
17 & Your mentor wrote a book on martial arts. Sometimes people seek you out to ask about its stranger passages. \\ \hline
18 & Someone you fought alongside in the military is now the mayor of a nearby town. \\ \hline
19 & You saved the lives of a family when their house burned down. They’re indebted to you, and their neighbors regard you as a hero. \\ \hline
20 & Your old trainer still expects you to come back and clean up after their classes; when you do, they occasionally share interesting rumors. \\ \hline

\end{tabularx}

\end{table*}

\subsection{Warrior Player Intrusions}\index{Types!Warrior!Player Intrusions}

You can spend 1 XP to use one of the following player intrusions, provided the situation is appropriate and the GM agrees.

\textbf{Perfect Setup}: You’re fighting at least three foes and each one is standing in exactly the right spot for you to use a move you trained in long ago, allowing you to attack all three as a single action. Make a separate attack roll for each foe. You remain limited by the amount of Effort you can apply on one action.

\textbf{Old Friend}: A comrade in arms from your past shows up unexpectedly and provides aid in whatever you’re doing. They are on a mission of their own and can’t stay longer than it takes to help out, chat for a while after, and perhaps share a quick meal.
Weapon Break: Your foe’s weapon has a weak spot. In the course of the combat, it quickly becomes damaged and moves two steps down the object damage track.

\subsection{First-tier Warrior}\index{Types!Warrior!1st Tier Warrior}

First-tier warriors have the following abilities:

\textbf{Effort}: Your Effort is 1.

\textbf{Physical Nature}: You have a Might Edge of 1 and a Speed Edge of 0, or you have a Might Edge of 0 and a Speed Edge of 1. Either way, you have an Intellect Edge of 0.

\textbf{Cypher Use}: You can bear two cyphers at a time.

\textbf{Weapons}: You become practiced with light, medium, and heavy weapons and suffer no penalty when using any kind of weapon. Enabler.

\textbf{Starting Equipment}: Appropriate clothing and two weapons of your choice, plus one expensive item, two moderately priced items, and up to four inexpensive items.

\textbf{Special Abilities}: Choose four of the abilities listed below. You can’t choose the same ability more than once unless its description says otherwise. The full description for each listed ability can be found in Abilities, which also has descriptions for flavor and focus abilities in a single vast catalog.

\begin{itemize}
\item Bash
\item Combat Prowess
\item Control the Field
\item Improved Edge
\item No Need for Weapons
\item Overwatch
\item Physical Skills
\item Practiced in Armor
\item Quick Throw
\item Swipe
\item Trained Without Armor
\end{itemize}

\subsection{Second-tier Warrior}\index{Types!Warrior!2nd Tier Warrior}

Choose two of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Crushing Blow
\item Hemorrhage
\item Reload
\item Skill With Attacks
\item Skill With Defense
\item Successive Attack
\end{itemize}


\subsection{Third-tier Warrior}\index{Types!Warrior!3rd Tier Warrior}

Choose three of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Deadly Aim
\item Energy Resistance
\item Experienced in Armor
\item Expert Cypher Use
\item Fury
\item Lunge
\item Reaction
\item Seize the Moment
\item Slice
\item Spray
\item Trick Shot
\item Vigilance
\end{itemize}

\subsection{Fourth-tier Warrior}\index{Types!Warrior!4th Tier Warrior}

Choose two of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Amazing Effort
\item Capable Warrior
\item Experienced Defender
\item Feint
\item Increased Effects
\item Momentum
\item Pry Open
\item Snipe
\item Tough As Nails
\end{itemize}

\subsection{Fifth-tier Warrior}\index{Types!Warrior!5th Tier Warrior}

Choose three of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Adroit Cypher Use
\item Arc Spray
\item Improved Success
\item Jump Attack
\item Mastery in Armor
\item Mastery With Attacks
\item Mastery With Defense
\item Parry
\end{itemize}

(Remember that at higher tiers, you can choose special abilities from lower tiers. This is sometimes the best way to ensure that you have exactly the character you want. This is particularly true with abilities that grant skills, which can usually be taken multiple times.)

\subsection{Sixth-tier Warrior}\index{Types!Warrior!6th Tier Warrior}

Choose two of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Again and Again
\item Finishing Blow
\item Magnificent Moment
\item Murderer
\item Spin Attack
\item Weapon and Body
\end{itemize}

\subsection{Warrior Example}\index{Types!Warrior!Warrior Example}

Ray wants to create a Warrior character for a modern campaign. He decides that the character is an ex-military fellow who is fast and strong. He puts 3 of his additional points into his Might Pool and 3 into his Speed Pool; his stat Pools are now Might 13, Speed 13, and Intellect 8. As a first-tier character, his Effort is 1, his Might Edge is 1, and his Speed Edge and Intellect Edge are both 0. His character is not particularly smart or charismatic.

He wants to use a large combat knife (a medium weapon that inflicts 4 points of damage) and a .357 Magnum (a heavy pistol that inflicts 6 points of damage but requires the use of both hands). Ray decides not to wear armor, as it’s not really appropriate to the setting, so for his first ability, he chooses Trained Without Armor so he eases Speed defense actions. For his second ability, he chooses Combat Prowess so he can inflict extra damage with his big knife. 

Ray wants to be fast as well as tough, so he selects Improved Edge. This gives him a Speed Edge of 1. He rounds out his character with Physical Skills and chooses swimming and running.

The Warrior can bear two cyphers. The GM decides that Ray’s first cypher is a pill that restores 6 points of Might when swallowed, and his second is a small, easily concealed grenade that explodes like a firebomb when thrown, inflicting 3 points of damage to all within immediate range. 

Ray still needs to choose a descriptor and a focus. Looking ahead to the descriptor rules, Ray chooses Strong, which increases his Might Pool to 17. He also becomes trained in jumping and breaking inanimate objects. (If he had chosen jumping as one of his physical skills, the Strong descriptor would have made him specialized in jumping instead of trained.) Being Strong also gives Ray an extra medium or heavy weapon. He chooses a baseball bat that he’ll use in a pinch. He keeps it in the trunk of his car.

For his focus, Ray chooses Masters Weaponry. This gives him yet another weapon of high quality. He chooses another combat knife and asks the GM if he could use it in his left hand—not to make attacks, but as a shield. This will ease his Speed defense rolls if he has both knives out (the “shield” counts as an asset). The GM agrees. During the game, Ray’s Warrior will be hard to hit—he is trained in Speed defense rolls, and his extra knife eases his defense rolls by another step.

Thanks to his focus, he also inflicts 1 additional point of damage with his chosen weapon. Now he inflicts 6 points of damage with his blade. Ray’s character is a deadly combatant, likely starting the game with a reputation as a knife fighter.

For his character arc, Ray chooses Defeat a Foe. That foe, he decides, is none other than someone in his company who was once a friend but went rogue.

\section{Adept}\index{Types!Adept}
\textbf{Fantasy/Fairy tale}: wizard, mage, sorcerer, cleric, druid, seer, diabolist, fey-touched

\textbf{Modern/Horror/Romance}: psychic, occultist, witch, practitioner, medium, fringe scientist

\textbf{Science fiction}: psion, psionicist, telepath, seeker, master, scanner, ESPer, abomination

\textbf{Superhero/Post-Apocalyptic}: mage, sorcerer, power-wielder, master, psion, telepath

You master powers or abilities outside the experience, understanding, and sometimes belief of others. They might be magic, psychic powers, mutant abilities, or just a wide variety of intricate devices, depending on the setting. (“Magic” here is a term used very loosely. It’s a catch-all for the kinds of wondrous, possibly supernatural things that your character can do that others cannot. It might actually be an expression of technological devices, channeling spirits, mutations, psionics, nanotechnology, or any number of other sources.)

\textbf{Individual Role}: Adepts are usually thoughtful, intelligent types. They often think carefully before acting and rely heavily on their supernatural abilities.

\textbf{Group Role}: Adepts are not powerful in straightforward combat, although they often wield abilities that provide excellent combat support, both offensively and defensively. They sometimes possess abilities that facilitate overcoming challenges. For example, if the group must get through a locked door, an Adept might be able to destroy it or teleport everyone to the other side. 

\textbf{Societal Role}: In settings where the supernatural is rare, strange, or feared, Adepts are likely rare and feared as well. They remain hidden, shadowy figures. When this is not the case, Adepts are more likely to be common and forthright. They might even take leadership roles.

\textbf{Advanced Adepts}: Even at low tiers, Adept powers are impressive. Higher-tier Adepts can accomplish amazing deeds that can reshape matter and the environment around them.

(Adepts are almost always emblematic of the paranormal or superhuman in some way—wizards, psychics, or something similar. If the game you’re playing has none of that, an Adept could be a charlatan mimicking such abilities with tricks and hidden devices, or a gadgeteer character with a “utility belt” full of oddments. Or a game like that might not have Adepts. That’s okay too.)

\subsection{Adept Player Intrusions}

When playing an Adept, you can spend 1 XP to use one of the following player intrusions, provided the situation is appropriate and the GM agrees.

\textbf{Advantageous Malfunction}: A device being used against you malfunctions. It might harm the user or one of their allies for a round, or activate a dramatic and distracting side effect for a few rounds.

\textbf{Convenient Idea:} A flash of insight provides you with a clear answer or suggests a course of action with regard to an urgent question, problem, or obstacle you’re facing.

\textbf{Inexplicably Unbroken}: An inactive, ruined, or presumed-destroyed device temporarily activates and performs a useful function relevant to the situation. This is enough to buy you some time for a better solution, alleviate a complication that was interfering with your abilities, or just get you one more use out of a depleted cypher or artifact.

\subsection{Adept Stat Pools}\index{Types!Adept!Stat Pools}

\begin{table}[H]
\centering
\caption{Adept Stat Pools}
\label{tab: Adept Stat Pools}

\begin{tabularx}{\linewidth}{| X | X |}
% \begin{tabularx}{\linewidth}{| p{0.5\linewidth} | p{0.5\linewidth} |}
\hline
\textbf{Stat} & \textbf{Pool Starting Value} \\
\hline
% Tutaj wpisujesz dane do tabeli
Might & 7 \\ \hline
Speed & 9 \\ \hline
Intelect & 12  \\ \hline

\end{tabularx}

\end{table}

\raggedright

You get 6 additional points to divide among your stat Pools however you wish.

\subsection{Adept Background Connection}\index{Types!Adept!Background Connection}

Your type helps determine the connection you have to the setting. Roll a d20 or choose from the following list to determine a specific fact about your background that provides a connection to the rest of the world. You can also create your own fact.

\begin{table*}
\centering

\caption{Adept Background Connection}
\label{tab:Adept Background Connection}

\begin{tabularx}{\textwidth}{| p{0.05\textwidth} | p{0.9\textwidth} |}
\hline
\textbf{d20} & \textbf{Background} \\
\hline
% Tutaj wpisujesz dane do tabeli
1 & You served as an apprentice for an Adept respected and feared by many people. Now you bear their mark. \\ \hline
2 & You studied in a school infamous for its dark, brooding instructors and graduates. \\ \hline
3 & You learned your abilities in the temple of an obscure god. Its priests and worshippers, although small in number, respect and admire your talents and potential.  \\ \hline
4 & While traveling alone, you saved the life of a powerful person. They remain indebted to you.  \\ \hline
5 & Your mother was a powerful Adept while she lived, helpful to many locals. They look upon you kindly, but they also expect much from you.  \\ \hline
6 & You owe money to a number of people and don’t have the funds to pay your debts. \\ \hline
7 & You failed disgracefully at your initial studies with your teacher and now proceed on your own. \\ \hline
8 & You learned your skills faster than your teachers had ever seen before. The powers that be took notice and are paying close attention.  \\ \hline
9 & You killed a well-known criminal in self-defense, earning the respect of many and the enmity of a dangerous few.  \\ \hline
10 & You trained as a Warrior, but your Adept predilections eventually led you down a different path. Your former comrades don’t understand you, but they respect you.  \\ \hline
11 & While studying to be an Adept, you worked as an assistant for a bank, making friends with the owner and the clientele. \\ \hline
12 & Your family owns a large vineyard nearby known to all for its fine wine and fair business dealings. \\ \hline
13 & You trained for a time with a group of influential Adepts, and they still look upon you with fondness. \\ \hline
14 & You worked the gardens in the palace of an influential noble or person of wealth. They wouldn’t remember you, but you made friends with their young daughter. \\ \hline
15 & An experiment you conducted in the past went horribly awry. The locals remember you as a dangerous and foolhardy individual. \\ \hline
16 & You hail from a distant place where you were well known and regarded, but people here treat you with suspicion. \\ \hline
17 & People you meet seem put off by the strange birthmark on your face. \\ \hline
18 & Your best friend is also an Adept. You and your friend share discoveries and secrets readily. \\ \hline
19 & You know a local merchant very well. Since you give them so much business, they offer you discounts and special treatment. \\ \hline
20 & You belong to a secretive social club that gathers monthly to drink and talk. \\ \hline

\end{tabularx}

\end{table*}

\subsection{First-tier Adept}\index{Types!Adept!1st Tier Adept}

First-tier Adepts have the following abilities:

\textbf{Effort}: Your Effort is 1. 

\textbf{Genius}: You have an Intellect Edge of 1, a Might Edge of 0, and a Speed Edge of 0. 

\textbf{Expert Cypher Use}: You can bear three cyphers at a time. 
Starting Equipment: Appropriate clothing, plus two expensive items, two moderately priced items, and up to four inexpensive items of your choice.

\textbf{Weapons}: You can use light weapons without penalty. You have an inability with medium weapons and heavy weapons; your attacks with medium and heavy weapons are hindered.

\textbf{Special Abilities}: Choose four of the abilities listed below. You can’t choose the same ability more than once unless its description says otherwise. The full description for each listed ability can be found in Abilities, which also has descriptions for flavor and focus abilities in a single vast catalog. (Adept abilities require at least one free hand unless the GM says otherwise.)

\begin{itemize}
\item Distortion
\item Erase Memories 
\item Far Step
\item Hedge Magic
\item Magic Training
\item Onslaught
\item Push
\item Resonance Field
\item Scan
\item Shatter
\item Ward
\end{itemize}

\subsection{Second-tier Adept}\index{Types!Adept!2nd Tier Adept}

Choose one of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Adaptation
\item Cutting Light
\item Hover
\item Mind Reading 
\item Retrieve Memories
\item Reveal
\item Stasis
\end{itemize}

\subsection{Third-tier Adept}\index{Types!Adept!3rd Tier Adept}

Choose two of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Adroit Cypher Use
\item Countermeasures
\item Energy Protection
\item Fire and Ice
\item Force Field Barrier
\item Sensor
\item Targeting Eye
\end{itemize}

\subsection{Fourth-tier Adept}\index{Types!Adept!4th Tier Adept}

Choose one of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Death Touch
\item Exile
\item Invisibility
\item Matter Cloud
\item Mind Control
\item Projection
\item Rapid Processing 
\item Regeneration
\item Reshape
\item Wormhole
\end{itemize}

\subsection{Fifth-tier Adept}\index{Types!Adept!5th Tier Adept}

Choose two of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Absorb Energy
\item Concussion
\item Conjuration
\item Create
\item Dust to Dust 
\item Knowing the Unknown
\item Master Cypher Use
\item Teleportation
\item True Senses
\end{itemize}

\subsection{Sixth-tier Adept}\index{Types!Adept!6th Tier Adept}

Choose one of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Control Weather
\item Earthquake
\item Move Mountains
\item Traverse the Worlds
\item Usurp Cypher
\end{itemize}

\subsection{Adept Example}\index{Types!Adept!Adept Example}

Jen wants to create an Adept—a sorcerer for a fantasy campaign. She decides to be somewhat well rounded, so she puts 2 of her additional points into each stat Pool, giving her a Might Pool of 9, a Speed Pool of 11, and an Intellect Pool of 14. Her Adept is smart and quick. She has an Intellect Edge of 1, a Might Edge of 0, and a Speed Edge of 0. As a first-tier character, her Effort is 1. As her initial abilities, she chooses Onslaught and Ward, giving her a strong offense and defense. She also chooses Magic Training and rounds out her character with Scan, which she hopes will be useful in gaining insight and information. For this character, Onslaught, Ward, and Scan are all spells she has mastered through years of training and study.

She can bear three cyphers. The GM gives her a potion that acts as a short-range teleporter, a small charm that restores 5 points to her Intellect Pool, and a fluid-filled flask that explodes like a fiery bomb. Jen’s sorcerer is skilled with light weapons, so she chooses a dagger. 
For her descriptor, Jen chooses Graceful, which adds 2 points to her Speed Pool, bringing it to 13. That descriptor means she is trained in balancing and anything requiring careful movements, physical performing arts, and Speed defense tasks. Perhaps she is a dancer. In fact, she begins to develop a backstory that involves graceful, lithe movements that she incorporates into her spells.

For her focus, she chooses Leads. This gives her training in social interactions, which again helps round her out—she’s good in all kinds of situations. Moreover, she has the Good Advice ability, which enables her to be a focal point of her group.

Her spells and focus abilities cost Intellect points to activate, so she’s glad to have a lot of points in her Intellect Pool. In addition, her Intellect Edge will help reduce those costs. If she uses her Onslaught force blast without applying Effort, it costs 0 Intellect points and deals 4 points of damage. Her Intellect Edge will allow her to save points to devote toward applying Effort for other purposes, perhaps to boost the accuracy of Onslaught.

For her character arc, Jen chooses Aid a Friend. She decides that when her sorcerer character was young, she had a magical mentor. That mentor was later taken prisoner by a demon, so her character is always looking for clues on how to find the demon and release her friend from bondage.

(GMs are always free to pre-select a type’s special abilities at a given tier to reinforce the setting. In the fantasy setting of Jen’s sorcerer, the GM might have said that all sorcerers (Adepts) start with Magic Training as one of their tier 1 abilities. This doesn’t make the character any less powerful or special, but it says something about her role in the world and expectations in the game.)

\section{Explorer}\index{Types!Explorer}

\textbf{Fantasy/Fairy tale}: Explorer, adventurer, delver, mystery seeker

\textbf{Modern/Horror/Romance}: athlete, explorer, adventurer, drifter, detective, scholar, spelunker, trailblazer, investigative reporter

\textbf{Science fiction}: Explorer, adventurer, wanderer, planetary specialist, xenobiologist

\textbf{Superhero/Post-Apocalyptic}: adventurer, crimefighter

You are a person of action and physical ability, fearlessly facing the unknown. You travel to strange, exotic, and dangerous places, and discover new things. This means you’re physical but also probably knowledgeable.

\textbf{Individual Role}: Although Explorers can be academics or well studied, they are first and foremost interested in action. They face grave dangers and terrible obstacles as a routine part of life. 

\textbf{Group Role}: Explorers sometimes work alone, but far more often they operate in teams with other characters. The Explorer frequently leads the way, blazing the trail. However, they’re also likely to stop and investigate anything intriguing they stumble upon. 

\textbf{Societal Role}: Not all Explorers are out traipsing through the wilderness or poking about an old ruin. Sometimes, an Explorer is a teacher, a scientist, a detective, or an investigative reporter. In any event, an Explorer bravely faces new challenges and gathers knowledge to share with others.

\textbf{Advanced Explorers}: Higher-tier Explorers gain more skills, some combat abilities, and a number of abilities that allow them to deal with danger. In short, they become more and more well-rounded, able to deal with any challenge.

\subsection{Explorer Stat Pools}\index{Type!Explorer!Stat Pools}

\begin{table}[H]
\centering
\caption{Explorer Stat Pools}
\label{tab: Explorer Stat Pools}

\begin{tabularx}{\linewidth}{| X | X |}
% \begin{tabularx}{\linewidth}{| p{0.5\linewidth} | p{0.5\linewidth} |}
\hline
\textbf{Stat} & \textbf{Pool Starting Value} \\
\hline
% Tutaj wpisujesz dane do tabeli
Might & 10 \\ \hline
Speed & 9 \\ \hline
Intelect & 9  \\ \hline

\end{tabularx}

\end{table}

\raggedright

You get 6 additional points to divide among your stat Pools however you wish.

\subsection{Explorer Background Connection}\index{Type!Explorer!Background Connection}

Your type helps determine the connection you have to the setting. Roll a d20 or choose from the following list to determine a specific fact about your background that provides a connection to the rest of the world. You can also create your own fact.

\begin{table*}
\centering

\caption{Explorer Background Connection}
\label{tab:Explorer Background Connection}

\begin{tabularx}{\textwidth}{| p{0.05\textwidth} | p{0.9\textwidth} |}
\hline
\textbf{d20} & \textbf{Background} \\
\hline
% Tutaj wpisujesz dane do tabeli
1 & You were a star high school athlete. You’re still in great shape, but those were the glory days, man. \\ \hline
2 & Your brother is the lead singer in a really popular band. \\ \hline
3 & You have made a number of discoveries in your explorations, but not all opportunities to capitalize on them have panned out yet.  \\ \hline
4 & You were a cop, but you gave it up after encountering corruption on the force.  \\ \hline
5 & Your parents were missionaries, so you spent much of your young life traveling to exotic places. \\ \hline
6 & You served in the military with honor. \\ \hline
7 & You received assistance from a secretive organization, which paid for your schooling. Now they seem to want a lot more from you. \\ \hline
8 & You went to a prestigious university on an athletic scholarship, but you excelled in class as well as on the field.  \\ \hline
9 & Your best friend from your youth is now an influential member of the government.  \\ \hline
10 & You used to be a teacher. Your students remember you fondly. \\ \hline
11 & You worked as a small-time criminal operative until you were caught and served some time in jail, after which you tried to go straight. \\ \hline
12 & Your greatest discovery to date was stolen by your arch-rival. \\ \hline
13 & You belong to an exclusive organization of Explorers whose existence is not widely known. \\ \hline
14 & You were kidnapped as a small child under mysterious circumstances, although you were recovered safely. The case still has some notoriety. \\ \hline
15 & When you were young, you were addicted to narcotics, and now you are a recovering addict. \\ \hline
16 & While exploring a remote location, you saw something strange you’ve never been able to explain. \\ \hline
17 & You own a small bar or restaurant. \\ \hline
18 & You published a book about some of your exploits and discoveries, and it has achieved some acclaim. \\ \hline
19 & Your sister owns a store and gives you a hefty discount. \\ \hline
20 & Your father is a high-ranking officer in the military with many connections. \\ \hline

\end{tabularx}

\end{table*}

\subsection{First-Tier Explorer}\index{Type!Explorer!1st Tier Explorer}

First-tier Explorers have the following abilities:

\textbf{Effort}: Your Effort is 1.

\textbf{Physical Nature}: You have a Might Edge of 1, a Speed Edge of 0, and an Intellect Edge of 0.

\textbf{Cypher Use}: You can bear two cyphers at a time.
Starting Equipment: Appropriate clothing and a weapon of your choice, plus two expensive items, two moderately priced items, and up to four inexpensive items.

\textbf{Weapons}: You can use light and medium weapons without penalty. You have an inability with heavy weapons; your attacks with heavy weapons are hindered.

\textbf{Special Abilities}: Choose four of the abilities listed below. You can’t choose the same ability more than once unless its description says otherwise. The full description for each listed ability can be found in Abilities, which also has descriptions for flavor and focus abilities in a single vast catalog.

\begin{itemize}
\item Block
\item Danger Sense
\item Decipher
\item Endurance
\item Find the Way
\item Fleet of Foot
\item Improved Edge
\item Knowledge Skills
\item Muscles of Iron
\item No Need for Weapons
\item Physical Skills
\item Practiced in Armor
\item Practiced With All Weapons
\item Surging Confidence
\item Trained Without Armor
\end{itemize}

\subsection{Second-Tier Explorer}\index{Type!Explorer!2nd Tier Explorer}

Choose four of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Curious
\item Danger Instinct
\item Enable Others
\item Escape
\item Eye for Detail 
\item Foil Danger
\item Hand to Eye
\item Investigative Skills
\item Quick Recovery
\item Range Increase
\item Skill With Defense
\item Stand Watch
\item Travel Skills
\item Wreck
\end{itemize}

\subsection{Third-Tier Explorer}\index{Type!Explorer!3rd Tier Explorer}

Choose three of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Controlled Fall
\item Experienced in Armor
\item Expert Cypher Use
\item Ignore the Pain
\item Obstacle Running
\item Resilience
\item Run and Fight
\item Seize the Moment
\item Skill With Attacks
\item Stone Breaker
\item Think Your Way Out
\item Trapfinder
\item Wrest From Chance
\end{itemize}

\subsection{Fourth-Tier Explorer}\index{Type!Explorer!4th Tier Explorer}

Choose two of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Capable Warrior
\item Expert Skill
\item Increased Effects
\item Read the Signs
\item Runner
\item Subtle Steps
\item Tough As Nails
\end{itemize}

\subsection{Fifth-Tier Explorer}\index{Type!Explorer!5th Tier Explorer}

Choose three of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Adroit Cypher Use
\item Free to Move
\item Group Friendship
\item Hard to Kill
\item Jump Attack
\item Mastery With Defense
\item Parry
\item Physically Gifted
\item Take Command
\item Vigilant
\end{itemize}

\subsection{Sixth-Tier Explorer}\index{Type!Explorer!6th Tier Explorer}

Choose three of the abilities listed below (or from a lower tier) to add to your repertoire. In addition, you can replace one of your lower-tier abilities with a different one from a lower tier.

\begin{itemize}
\item Again and Again
\item Inspire Coordinated Actions
\item Mastery in Armor
\item Mastery With Attacks
\item Negate Danger
\item Share Defense
\item Spin Attack
\item Wild Vitality
\end{itemize}

\subsection{Explorer Example}\index{Type!Explorer!Explorer Example}

Sam decides to create an Explorer character for a science fiction campaign. This character will be a hardy soul who explores alien worlds. They put 3 additional points into their Might Pool, 2 into their Speed Pool, and 1 into their Intellect Pool; their stat Pools are now Might 13, Speed 11, and Intellect 10. As a first-tier character, their Effort is 1, their Might Edge is 1, and their Speed Edge and Intellect Edge are 0. Their character is fairly well-rounded so far.

Sam immediately leaps in and starts choosing abilities. They pick Danger Sense and Surging Confidence, thinking those abilities will be generally useful. They also choose Practiced in Armor, reasoning that the character wears high-tech medium armor when exploring. Last, they choose Knowledge Skills and select geology and biology to help during interplanetary explorations. 

Sam’s Explorer can bear two cyphers, which in this setting involve nanotechnology. The GM decides that one is a nanite injector that grants a +1 bonus to Might Edge when used, and the other is a device that can create one simple handheld object the user wishes. 

Sam’s Explorer is not really geared toward fighting, but sometimes the universe is a dangerous place, so they note that they’re carrying a medium blaster as well.
Sam still needs a descriptor and a focus. Looking to the Descriptor chapter, they choose Hardy, which increases their Might Pool to 17. They also heal more quickly and can operate better when injured. They’re trained in Might defense but have an inability with initiative; however, it’s effectively canceled out by their Danger Sense (and vice versa). Sam could go back and select something else instead of Danger Sense, but they like it and decide to keep it. Overall, the descriptor ends up making the character tough but a little slow.

For their focus, Sam chooses Explores Dark Places (in this case, weird ruins of alien civilizations). This gives the character a bunch of additional skills: searching, listening, climbing, balancing, and jumping. They’re quite the capable Explorer.

For their character arc, Sam chooses Enterprise. Exploring alien places sometimes turns up strange relics, and Sam figures they might be able to set up a service to reliably transport these items to responsible third parties, rather than allow them to fall into the hands of pirates and rich private collectors. For a small fee, of course.

\section{Speaker}\index{Types!Speaker}

\textbf{Fantasy/Fairy tale}: bard, speaker, skald, emissary, priest, advocate

\textbf{Modern/Horror/Romance}: diplomat, charmer, face, spinner, manipulator, minister, mediator, lawyer

\textbf{Science fiction}: diplomat, empath, glam, consul, legate

\textbf{Superhero/Post-Apocalyptic}: charmer, mesmerist, puppet master

You’re good with words and good with people. You talk your way past challenges and out of jams, and you get people to do what you want.

\textbf{Individual Role}: Speakers are smart and charismatic. They like people and, more important, they understand them. This helps speakers get others to do what needs to be done. 

\textbf{Group Role}: The Speaker is often the face of the group, serving as the person who speaks for all and negotiates with others. Combat and action are not a Speaker’s strong suits, so other characters sometimes have to defend the Speaker in times of danger.

\textbf{Societal Role}: Speakers are frequently political or religious leaders. Just as often, however, they are con artists or criminals. 
Advanced Speakers: Higher-tier speakers use their abilities to control and manipulate people as well as aid and nurture their friends. They can talk their way out of danger and even use their words as weapons.

\subsection{Speaker Player Intrusions}\index{Types!Speaker!Player Intrusions}

When playing a Speaker, you can spend 1 XP to use one of the following player intrusions, provided the situation is appropriate and the GM agrees.

\textbf{Friendly NPC}: An NPC you don’t know, someone you don’t know that well, or someone you know but who hasn’t been particularly friendly in the past chooses to help you, though doesn’t necessarily explain why. Maybe they’ll ask you for a favor in return afterward, depending on how much trouble they go to.

\textbf{Perfect Suggestion}: A follower or other already-friendly NPC suggests a course of action with regard to an urgent question, problem, or obstacle you’re facing.

\textbf{Unexpected Gift}: An NPC hands you a physical gift you were not expecting, one that helps put the situation at ease if things seem strained, or provides you with a new insight for understanding the context of the situation if there’s something you’re failing to understand or grasp.

\subsection{Speaker Stat Pools}\index{Types!Speaker!Stat Pools}

\begin{table}[H]
\centering
\caption{Speaker Stat Pools}
\label{tab: Speaker Stat Pools}

\begin{tabularx}{\linewidth}{| X | X |}
% \begin{tabularx}{\linewidth}{| p{0.5\linewidth} | p{0.5\linewidth} |}
\hline
\textbf{Stat} & \textbf{Pool Starting Value} \\
\hline
% Tutaj wpisujesz dane do tabeli
Might & 8 \\ \hline
Speed & 9 \\ \hline
Intelect & 11  \\ \hline

\end{tabularx}

\end{table}

\raggedright

You get 6 additional points to divide among your stat Pools however you wish. 

\subsection{Speaker Background Connection}\index{Types!Speaker!Background Connection}

Your type helps determine the connection you have to the setting. Roll a d20 or choose from the following list to determine a specific fact about your background that provides a connection to the rest of the world. You can also create your own fact.

\end{multicols}
